---
title: "COVID plots"
description: |
  Plotting some data on COVID that made me rethink things
author:
  - name: Roy Emanuel
    orcid_id: 0000-0003-2935-2651
    url: {}
categories:
  - data
  - COVID
date: 02-05-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("d:/COVID/deathbyweek.r")
library(pander)
library(captioner)
table_nums <- captioner(prefix = "Table")
fig_nums <- captioner(prefix = "Figure")
```

# Background

I have been building my own plots for COVID-19 fatality since April of
last year. These plots have been my little pet project for me to view
on my own. In general the swells and bumps followed the narrative told
as far as increasing deaths and decreasing deaths. I did not go so far
as to plot hospital bed usage and stuff, and I'm not going there. I
have just three plots based on age that made me realize I had false
sense of the impact of the virus. 

 Here are the
numbers as of 4 FEB 2021 mixed with the headline data from the
[Washington Post COVID-19
Tracker](https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/):

```{r usTotal}
usTotal <- tibble(`Total reported US cases` = "26,611,000",
               `Total US deaths` = "454,000")

usTotal %>% pander()
```
`r table_nums(name = "UStotal", caption = "Total US Values as reported
by Washington Post COVID-19 Tracker")`

# Long-Term Care Facilities

I decided to look into the age breakdowns for deaths after seeing the
Atlantic's
[Long-Term Care COVID
Tracker](https://covidtracking.com/nursing-homes-long-term-care-facilities). 

You can deep dive, but the top line tells the story.

```{r ltcf}


ltcf <- tibble(`Total LTCF cases` = "1,234,012",
               `Total LTCF deaths` = "153,159",
               `Total number of LTCF affected` = "31,517")


ltcf %>% pander()
```
`r table_nums(name = "LTCFtotal", caption = "Total Long Term Care Facilities (LTCF) data as reported by the Atlantic COVID-19")`

I find this incredible and not talked about. Sure, the media points
out that older people and those with co-morbidities (other health
problems) have a higher fatality rate. But not *this*. **OVER ONE
THIRD of COVID-19 deaths occur in Long-Term Care Facilities!** My gut
feeling from the news and social media was that maybe an extra 10-20%
of deaths were in the 70 and older group. Not this.


# United States by Age Groups

The findings from the  [Long-Term Care COVID
Tracker](https://covidtracking.com/nursing-homes-long-term-care-facilities)
made me dig deeper to find the percentage of COVID deaths by age, the
number of cases by age, and the number of cases per death by
age. `r fig_nums(name = "AgeDist", caption = "Percentage of Total US COVID-19 Deaths Suffered by Age Groups. Data from CDC", display = "cite")` shows the
percent of total COVID-19 deaths suffered by each age group.

```{r ageDist, fig.height = 6}
US_death_by_age +
    scale_y_continuous(name = "Percentage of Total COVID-19 Deaths") +
    theme(legend.position = "none")
```

`r fig_nums(name = "AgeDist")`

Less than 3% of US deaths have been people below the age of 43. Well
under 1% of the deaths have been young adults and children (<24 years
old). Over 80% of deaths are from the 65+ age groups. All data came
from the Center for Disease Control data website [United States
COVID-19 Cases and Deaths by State over
Time](https://data.cdc.gov/Case-Surveillance/United-States-COVID-19-Cases-and-Deaths-by-State-o/9mfq-cb36). 


# California by Age Groups

The California Public Health Department has a straightforward website
with a table of data that is much easier to read than their fancy
Tableau web pages. The data has number of cases, number of deaths and
percentage of the California population in [COVID-19 Cases by Age
Group](https://www.cdph.ca.gov/Programs/CID/DCDC/Pages/COVID-19/COVID-19-Cases-by-Age-Group.aspx). Data
is from 5 FEB 2021. 

`r fig_nums(name = "CAperc", caption = "Age Aggregated COVID-19 Case and Death
Percentage compared to Population Percentage", display = "cite")` 
shows how the case and death percentage compare to the overall
population percentage of each age group.

```{r CAperc, fig.height = 8}
CA_percent_caseDeathPop +
    theme(legend.position = "bottom")
```

`r fig_nums(name = "CAperc")`

Now the same plots but using the count data rather than relative
percentages. 

```{r CAcount, fig.height = 8}
CA_count_caseDeathPop +
    theme(legend.position = "bottom")
```

The final plot took a small amount of math on my part. I wanted to see
how many cases per fatality there are for each age group. This gives
an idea of if you have a group with a  reported positive test, how
many of them died. This is simplified. It does not take into account
active cases that may become deaths or unreported or asymptomatic
cases. My personal belief is that these numbers are pessimistic
because the number of asymptomatic or unreported cases is
higher than the amount of current cases that will result in death. Also,
they are more pessimistic for the younger groups than the older groups
because I believe younger groups are less likely to get tested and are
more likely to have asymptomatic COVID-19.

```{r CAcaseperdeath, fig.height = 8}
case_per_death_by_age
```

<!-- # Opinion or So What? -->

<!-- This poor messaging hurts the government response to COVID-19 in two -->
<!-- ways. First is the risk to the elderly, I believe is severely under -->


